### 2021-11-14 text/outline/draft.tex
complete draft of all math content

\documentclass[12pt]{article}
 \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{array}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother
\setcounter{tocdepth}{1}
\usepackage{xcolor}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\usepackage{float}
\allowdisplaybreaks
\usepackage[caption = false]{subfig}
\usepackage{/Users/timbarry/optionFiles/mymacros}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}
\noindent
Tim, Gene, Kathryn
\begin{center}
\textbf{CRISPR genome editing, single-cell sequencing, and exponential family measurement error models}
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. In this work we study single-cell CRISPR screens from a statistical perspective. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (“generalized linear model with errors-in-variables”), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents

\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

Our work makes two main contributions. First, we study the thresholding method from empirical and theoretical perspectives. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model. \blue{Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. In doing so, GLM-EIV implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}. These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders. In this work we tackle a set of analysis challenges that are complimentary to the challenges addressed by SCEPTRE. Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{an unattainable objective within the nonparameteric hypothesis testing framework of SCEPTRE.}

@@

\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design. For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed. Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene. \textbf{b,} DAG representing all variables in the system. The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as nuisance variables, also impacting gene and gRNA expression. The target of estimation is the effect of the perturbation on gene expression. \textbf{c,} Schematic illustrating the ``background read'' phenomenon. Due to errors in the sequencing process, unperturbed cells exhibit a nonzero gRNA count distribution (bottom). The target of estimation is the change in mean gene expression in response to the perturbation (top). \textbf{d}, Example data on four cells for a given perturbation-gene pair. Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
	\label{analysis_challenges}
\end{figure}

@@

\subsubsection*{Log likelihood and model properties}
We derive the log-likelihood of the GLM-EIV model. We conduct estimation and inference \textit{conditional} on the library sizes and technical factors $l^m_i, l^g_i,$ and $z_i$; therefore, we treat these quantities as fixed constants. We assume that the gene expression $m_i$ and gRNA expression $g_i$ are \textit{conditionally independent} given the perturbation $p_i$. The joint density $f$ of $(m_i, g_i, p_i)$ given $\theta$ is
\begin{equation}\label{full_density}
f(m_i,g_i,p_i; \theta) = f_m(m_i | p_i) f_g(g_i | p_i) f_p(p_i) \\ = \pi^{p_i}(1-\pi)^{1-p_i} f_m(m_i; \eta^m_i) f_g(g_i; \eta^g).
\end{equation}
The log-likelihood is
\begin{multline}\label{full_log_lik}
\mathcal{L}(\theta; m, g, p) = \sum_{i=1}^n \log\left( \pi^{p_i}(1-\pi)^{1-p_i} \right) \\ + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right). \end{multline}
Integrating over the unobserved variable $p_i$, we can write the marginal density $f$ of $(m_i, g_i)$ as
\begin{equation}\label{marginal_density}
f(m_i, g_i; \theta) = (1-\pi) f(m_i; \eta^m_i(0)) f(g_i; \eta^g_i(0)) + \pi f(m_i; \eta^m_i(1)) f(g_i; \eta^g_i(1)).
\end{equation}
Finally, the marginal log-likelihood is
\begin{multline}\label{marginal_log_lik}
\mathcal{L}(\theta; m, g) \\ = \sum_{i=1}^n \log\left[ (1-\pi) f(m_i; \eta^m_i(0)) f(g_i; \eta^g_i(0)) + \pi f(m_i; \eta^m_i(1)) f(g_i; \eta^g_i(1)) \right]
\end{multline}

@@

After fitting the model, we perform inference on the estimated parameters. The easiest approach, given the complexity of the log likelihood, would be to run a parametric bootstrap on the fitted model. This strategy, however, is prohibitively slow, as the data are large and we use an EM algorithm to fit the model. Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982} (see appendix). Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data.

@@

Leveraging these technologies, we develop a scalable and efficient pipeline for deploying GLM-EIV (Algorithm \ref{algo:at_scale}). First, we run a round of ``precomputations'' on all $d_g$ genes and $d_p$ perturbations. The precomputations involve regressing the gene expressions (or gRNA counts) onto the technical factors, thereby ``factoring out'' lines 2-7 from Algorithm \ref{algo:pilot_estimates}. Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the set of pilot parameters (Algorithm \ref{algo:pilot_estimates}, lines 8 onward), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference. The three loops in Algorithm \ref{algo:at_scale} are embarassingly parallel and therefore can be massively parallelized.

@@

\subsection{Zero-inflated model}

@@

$$\left[\eta^g_i(k)\right]^{(t)} :=  h_m\left( \langle \tilde{x}_i(k) , \beta_g^{(t)} \rangle + o^g_i \right).$$

@@

T^{(t)}_i(1) = \P(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)}) \\ = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} \\ = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
where we set $$q_i^{(t)} := \log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right).$$

@@

\subsection{Inference}
We derive the asymptotic observed information matrix of the GLM-EIV log likelihood, enabling us to perform inference on the parameters. First, we define some notation. For $i \in \{1, \dots, n\}$, $j \in \{0, 1\}$, and $\theta = (\pi, \beta_m, \beta_g),$ let $T^\theta_i(j)$ be defined by
$$T^\theta_i(j) = \P_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Also, define the $\R^n$ vectors $T^\theta(0)$ and $T(1)$ by
$$T^\theta(0) := [T^\theta_1(0), T^\theta_2(0), \dots, T^\theta_n(0)]$$ and 
$$T^\theta(1) := [T^\theta_1(1), T^\theta_2(1), \dots, T^\theta_n(1)].$$

Next, define the diagonal $n \times n$ matrices $\Delta^m$, $[\Delta']^m$, $V^m$, and $H^m$ by
$$
\begin{cases}
	\Delta^m = \textrm{diag} \{ h_m'(l_1^m), \dots, h_m'( l_n^m ) \} \\
	[\Delta']^m = \textrm{diag} \{ h_m''(l_1^m), \dots, h_m''( l_n^m) \} \\
	V^m = \textrm{diag} \{ \psi''_m( \eta^m_1), \dots, \psi''_m( \eta^m_n) \} \\
	H^m = \textrm{diag} \{ m_1 - \mu_1^m, \dots, m_n - \mu_n^m\}.
\end{cases} 
$$ Define the $n \times n$ matrices $\Delta^g, [\Delta']^g, V^g,$ and $H^g$ analogously. These matrices are \textit{unobserved}, as they depend on $\{p_1, \dots, p_n\}$.

@@

\section{Zero-inflated model}
\subsection{Relationship to background real model}
\subsection{Estimation}
\subsection{Inference}

\section{Statistical accelerations}
A key step in the algorithm for computing the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) is to fit a weighted, no-intercept, univariate GLM with nonzero offset terms and a binary predictor variable (line 20). We derive an analytic formula for the MLE of this GLM for three important pairs of response distributions and link functions: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link. The GLM that we seek to estimate has responses $[m,m]^T$, predictors $[\underbrace{0, \dots, 0}_\textrm{n}, \underbrace{1, \dots, 1}_\textrm{n}]$, offsets $[\hat{f}^m, \hat{f}^m],$ and weights $w = [T_1(0), \dots, T_n(0), T_1(1), \dots, T_n(1)]^T.$ Throughout, $C$ denotes a universal constant. The log likelihood of this GLM is
\begin{multline*}
$$\mathcal{L}(\beta_1; m) = \sum_{i=1}^n T_i(0) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\hat{f}^m_i)) \\ = \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + C.$$
\end{multline*}
Thus, finding the MLE $\hat{\beta_1}$ is equivalent to estimating a GLM with intercept $\beta_1$, offsets $\hat{f}^m$, weights $T_i(1)$, and \textit{no} covariate terms. We term such a GLM a \textit{intercept-plus-offset} model. Below, we study intercept-plus-offset models in generality.

>>>>

\documentclass[12pt]{article}
 \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{array}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother
\setcounter{tocdepth}{2}
\usepackage{xcolor}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\usepackage{float}
\allowdisplaybreaks
\usepackage[caption = false]{subfig}
\usepackage{/Users/timbarry/optionFiles/mymacros}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}
\noindent
Tim, Gene, Kathryn
\begin{center} 
\textbf{A new class of exponential family measurement error models, with application to single-cell CRISPR screens}
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. In this work we study single-cell CRISPR screens from a statistical perspective. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (“generalized linear model with errors-in-variables”), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

Our work makes two main contributions. First, we conduct a careful study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model. \blue{Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. In doing so, GLM-EIV implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. % Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}. These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{an objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.}

@@

\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design. For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed. Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene. \textbf{b,} DAG representing all variables in the system. The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression. The target of estimation is the effect of the perturbation on gene expression. \textbf{c,} Schematic illustrating the ``background read'' phenomenon. Due to errors in the sequencing process, unperturbed cells exhibit a nonzero gRNA count distribution (bottom). The target of estimation is the change in mean gene expression in response to the perturbation (top). \textbf{d}, Example data on four cells for a given perturbation-gene pair. Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
	\label{analysis_challenges}
\end{figure}

@@

\subsubsection*{Log likelihood and model properties}
We derive the log-likelihood of the GLM-EIV model. We conduct estimation and inference \textit{conditional} on the library sizes and technical factors $l^m_i, l^g_i,$ and $z_i$; therefore, we treat these quantities as fixed constants. We assume that the gene expression $m_i$ and gRNA expression $g_i$ are \textit{conditionally independent} given the perturbation $p_i$. The joint density $f$ of $(m_i, g_i, p_i)$ given $\theta$ is
\begin{equation}\label{full_density}
f(m_i,g_i,p_i; \theta) = f_m(m_i | p_i) f_g(g_i | p_i) f_p(p_i) \\ = \pi^{p_i}(1-\pi)^{1-p_i} f_m(m_i; \eta^m_i) f_g(g_i; \eta^g).
\end{equation}
The log-likelihood is
\begin{multline}\label{full_log_lik}
\mathcal{L}(\theta; m, g, p) = \sum_{i=1}^n \log\left( \pi^{p_i}(1-\pi)^{1-p_i} \right) \\ + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right). \end{multline}
Integrating over the unobserved variable $p_i$, we can write the marginal density $f$ of $(m_i, g_i)$ as
\begin{equation}\label{marginal_density}
f(m_i, g_i; \theta) = (1-\pi) f_m(m_i; \eta^m_i(0)) f_g(g_i; \eta^g_i(0)) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i; \eta^g_i(1)).
\end{equation}
Finally, the marginal log-likelihood is
\begin{multline}\label{marginal_log_lik}
\mathcal{L}(\theta; m, g) \\ = \sum_{i=1}^n \log\left[(1-\pi) f_m(m_i; \eta^m_i(0)) f_g(g_i; \eta^g_i(0)) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i; \eta^g_i(1)) \right]
\end{multline}

@@

After fitting the model, we perform inference on the estimated parameters. The easiest approach, given the complexity of the log likelihood, would be to run a parametric bootstrap on the fitted model. This strategy, however, is prohibitively slow, as the data are large and we use an EM algorithm to fit the model. Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982} (see appendix). Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data. The derivation is fairly involved; simulation studies (subsequent section) confirm the correctness of the results.

@@

Leveraging these technologies, we develop a scalable and efficient pipeline for GLM-EIV (Algorithm \ref{algo:at_scale}). First, we run a round of ``precomputations'' on all $d_g$ genes and $d_p$ perturbations. The precomputations involve regressing the gene expressions (or gRNA counts) onto the technical factors, thereby ``factoring out'' lines 2-7 from Algorithm \ref{algo:pilot_estimates}. Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the complete set of pilot parameters (Algorithm \ref{algo:pilot_estimates}, lines 8 onward), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference. The three loops in Algorithm \ref{algo:at_scale} are embarassingly parallel and therefore can be massively parallelized.

@@

\subsection{Zero-inflated model}
The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes. This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}. However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero. We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}. We derive efficient methods for estimation and inference in this model similar to those outlined above. However, given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard (or ``background read'') GLM-EIV model for the remainder of the main text.

@@

$$\left[\eta^g_i(k)\right]^{(t)} :=  h_g\left( \langle \tilde{x}_i(k) , \beta_g^{(t)} \rangle + o^g_i \right).$$

@@

\begin{multline}\label{e_step_1}
T^{(t)}_i(1) = \P(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)})  = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} \\ = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
where we set 
\begin{equation}\label{e_step_2}
q_i^{(t)} := \log\left(\frac{\pi^{(t)}(0) \alpha_i^{(t)}(0)}{\pi^{(t)}(1) \alpha_i^{(t)}(1)}\right).

@@

\subsection{Inference}
We derive the asymptotic observed information matrix of the GLM-EIV log likelihood, enabling us to perform inference on the parameters. First, we define some notation. For $i \in \{1, \dots, n\}$, $j \in \{0, 1\}$, and $\theta = (\pi, \beta_m, \beta_g),$ let $T^\theta_i(j)$ be defined by
$$T^\theta_i(j) = \P_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Let the $n \times n$ matrix $T^\theta(j)$ be given by $$T^\theta(j) = \textrm{diag}\left\{T^\theta_1(j), \dots, T^\theta_n(j)\right\}.$$
Next, define the diagonal $n \times n$ matrices $\Delta^m$, $[\Delta']^m$, $V^m$, and $H^m$ by
$$
\begin{cases}
	\Delta^m = \textrm{diag} \{h_m'(l_1^m), \dots, h_m'(l_n^m)\} \\
	[\Delta']^m = \textrm{diag} \{h_m''(l_1^m), \dots, h_m''(l_n^m) \} \\
	V^m = \textrm{diag} \{ \psi''_m( \eta^m_1), \dots, \psi''_m( \eta^m_n) \} \\
	H^m = \textrm{diag} \{ m_1 - \mu_1^m, \dots, m_n - \mu_n^m\}.
\end{cases} 
$$ Define the $n \times n$ matrices $\Delta^g, [\Delta']^g, V^g,$ and $H^g$ analogously. These matrices are \textit{unobserved}, as they depend on $\{p_1, \dots, p_n\}$.

@@

\section{Zero-inflated model}\label{sec:zero_inf_model}
In this section we introduce the ``zero-inflated'' GLM-EIV model. The zero-inflated GLM-EIV model is appropriate to use when the unperturbed cells do not transcribe \textit{any} gRNA molecules (i.e., when there are no background reads). Let $x_i = [1, z_i]^T \in \R^{d-1}$ be the vector of observed covariates, including an intercept term. ($x_i$ is the same as $\tilde{x}_i$, but with the perturbation indicator $p_i$ removed.) Let $\beta_{g,z} = [\beta^g_0, \gamma_g] \in \R^{d-1}$ be an unknown coefficient vector. ($\beta_{g,z}$ is the same as $\beta_g$, but with the perturbation effect $\beta^g_1$ removed). Let the linear component $l^{g,z}_i$, mean $\mu^{g,z}_i$, and canonical parameter $\eta^{g,z}_i$ of gRNA count distribution of the $i$th cell be given by
$$\begin{cases}
l^{g,z}_i = \langle x_i, \beta_{g,z} \rangle + o^g_i \\
r_g(\mu^{g,z}_i) = l^{g,z}_i \\
\eta^{g,z}_i = ([\psi'_g]^{-1} \circ r^{-1}_g)(l^{g,z}_i) := h_g(l^{g,z}_i).
\end{cases}
$$
The density $f_{g,z}$ of gRNA counts in the zero-inflated model is as follows:
$$f_{g,z}(g_i; \eta^{g,z}_i, p_i) = [f_g(g_i;\eta^{g,z}_i)]^{p_i} \mathbb{I}(g_i = 0)^{1-p_i}.$$
In other words, when the cell is \textit{perturbed} (i.e., $p_i = 1$), the zero-inflated density $f_{g,z}$ coincides with the background-read density $f_g$; by contrast, when the cell is \textit{unperturbed} (i.e., $p_i = 0$), the zero-inflated density $f_{g,z}$ is a point mass at zero. The gene expression density $f_m$ and perturbation indicator density $f_p$ are the same across the background read and zero-inflated models. We assume that the gene expression $m_i$ and gRNA count $g_i$ are conditionally independent given the perturbation indicator $p_i$. The joint density $f_z$ of $(m_i, p_i, z_i)$ is
\begin{multline*}
f_z(m_i, g_i, p_i) = f_m(m_i | p_i) f_{g,z}(g_i | p_i) f_p(p_i) \\ = \pi^{p_i} (1-\pi)^{1-p_i} f_m(m_i; \eta^m_i)[f_g(g_i; \eta_i^{g,z})]^{p_i} \mathbb{I}(g_i = 0)^{1-p_i}.
\end{multline*}
The log-likelihood $\mathcal{L}_z$ is
\begin{multline*}
\mathcal{L}_z(\theta; m, g, p) = \sum_{i=1}^n \log \left[ \pi^{p_i} (1-\pi)^{1-p_i} \right] + \sum_{i=1}^n \log\left[f_m(m_i; \eta^m_i) \right] \\ + \sum_{i=1}^n p_i \log \left[f_{g}(g_i; \eta_i^{g,z}) \right] + \sum_{i=1}^n (1 - p_i)\log\left[ \mathbb{I}(g_i = 0) \right],
\end{multline*}
where $\theta = [\pi, \beta_m, \beta_{g,z}]$ is the vector of unknown parameters.
Integrating over the unobserved variable $p_i$, the marginal density $f_z$ of $(m_i, g_i)$ is
$$f_z(m_i, g_i; \theta) = (1-\pi) f_m(m_i;\eta^m_i(0)) \mathbb{I}(g_i = 0) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i;\eta^{g,z}_i).$$ Finally, the marginal log-likelihood is
$$
\mathcal{L}_z(\theta; m_i, g_i) = \sum_{i=1}^n \log\left[ (1-\pi) f_m(m_i;\eta^m_i(0)) \mathbb{I}(g_i = 0) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i;\eta^{g,z}_i) \right].
$$

\subsection{Estimation}
To estimate the parameters of the zero-inflated GLM-EIV model, we use an EM algorithm similar to Algorithm \ref{algo:em_full} but with two changes. First, we use a different formula for the $i$th membership probability at the $t$-th step of the algorithm $T^{(t)}_i(1)$. (We use $T^{(t)}_i(1)$ to denote the $i$th membership probability in \textit{both} the background read and zero inflated cases; the difference should be clear from context.) Let $\theta^{(t)} = (\pi^{(t)}, \beta^{(t)}_m, \beta^{(t)}_{g,z})$ be the parameter estimate at the $t$-th iteration of the algorithm. Arguing in a manner similar to the background read case, we have that $$T^{(t)}_i(1) = \frac{1}{ \exp(q_i^{\left(t,z\right)}) + 1},$$ where
$$q_i^{(t,z)} = \log \left(\frac{(1 - \pi^{(t)}) \P(M_i = m_i | P_i = 0, \theta^{(t)}) \P(G_i = g_i | P_i = 0, \theta^{(t)})}{(\pi^{(t)}) \P(M_i = m_i | P_i = 1, \theta^{(t)}) \P(G_i = g_i | P_i = 1, \theta^{(t)})}\right).$$
The expression for $q^{(t,z)}_i$ is
\begin{multline*}
q^{(t,z)}_i = \log\left[ 1 - \pi^{(t)} \right] + \log\left[ f_m\left(m_i; \left[ \eta^m_i(0) \right]^{(t)}\right) \right] + \log\left[ \mathbb{I}(g_i = 0) \right] \\ - \log\left[ \pi^{(t)} \right] - \log\left[ f_m\left(m_i; \left[ \eta^m_i(1) \right]^{(t)}\right) \right] - \log\left[ f_g\left(g_i; \left[\eta^{g,z}_i \right]^{(t)}\right) \right],
\end{multline*}
where $[\eta^{g,z}_i]^{(t)} = h_g( \langle x_i, \beta^{(t)}_{g,z} \rangle + o^g_i).$ Notice that if $g_i \geq 1$, then $T^{(t)}_i(1) = 1.$ This comports with our intuition that a nonzero gRNA count indicates the presence of a perturbation.

Next, we consider the M step of the EM algorithm, which is similar to the background read case but slightly different. Define $Q_z(\theta | \theta^{(t)}) = \E_{(P | M = m, G = g, \theta^{(t)})} \left[ \mathcal{L}_z(\theta; m, g, p) \right].$ We have that
\begin{multline}\label{q_funct_zero_inf}
Q_z(\theta | \theta^{(t)}) = \sum_{i=1}^n \left[ T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right] + \sum_{i=1}^n \sum_{j=0}^1 T^{(t)}_i(j) \log \left[ f_m( m_i; \eta_i^m(j)) \right] \\ + \sum_{i=1}^n T_i^{(t)}(1) \left[ \log( f_g(g_i; \eta_i^{g, z}))\right] + C.
\end{multline}
The three terms of (\ref{q_funct_zero_inf}) are functions of $\pi$, $\beta_m$, and $\beta_{g,z}$, respectively. The maximizer $\pi^{(t)}$ and $\beta_m^{(t+1)}$ of the first and second term are the same as in the background read case. The maximizer $\beta^{(t+1)}_{g,z}$ of the third term is the maximizer of the GLM with exponential family density $f_g$, link function $r_g$, responses $g$, weights $T^{(t)}(1)$, design matrix $X$, offsets $o^g$.
